State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-17 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
edits to results section

Given the basis function separation, the basis function width parameter was set to $\sigma_{\phi}^2=2.5$~mm. This allowed a smooth representation of the field by the basis function reconstruction up to a spatial frequency $\mathbf{\nu_{c\phi}} \approx 0.12$~Hz, as described by equation~\ref{eq:CutoffFromBasisFuncWidth}. The band-limiting effect is due to the spatial frequency response of the basis functions as seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
\begin{figure} 
	\begin{center}
		\subfigure[][]{\label{fig:ObservationFreqAnal}\includegraphics{./Graph/ObservationFrequencyResponse.pdf}}\\	
		\subfigure[][]{\label{fig:SensorFunctionFreqResponse}\includegraphics{./Graph/SensorChar.pdf}}
		\subfigure[][]{\label{fig:BasisFunctionFreqResponse}\includegraphics{./Graph/BasisChar.pdf}}
	\end{center}
	\caption{\dean{Can we use $\nu_{cy}$ and $\nu_{c\phi}$ in the labels on the graphs?} Spatial frequency analysis. The dashed line shows the cut-off frequency (-3 dB point). \textbf{a.} Two-dimensional spatial frequency in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the 1-dimensional representation of the spectral characteristics. \textbf{b.} The spatial frequency response of a 1-dimensional model sensor. \textbf{c.} The spatial frequency response of a 1-dimensional neural field basis function, illustrating a narrower band width than the sensors.} 
\end{figure}

\subsection{State and Parameter Estimates} 
\label{sec:state_and_param_results}
To demonstrate the performance of the parameter and state estimation, a Monte Carlo approach was employed where $100$ realisations of the data were generated. Each realisation consisted of $500$~ms of data (sampled at $1$~kHz) and the estimation was applied to the final $400$~ms, allowing the model's initial transients to die out. The initial state and parameters were unknown to the estimator. Eleven iterations of the estimation algorithm were used to ensure the parameters  converge.

The distribution of the estimated parameters is shown in Figure~\ref{fig:Parameters}. The mean parameter estimates and the corresponding standard deviations are given in Table~\ref{tab:Parameters estimates}. The estimates for $\xi$, where $\xi=1-\zeta T_s $ and $\zeta$ is the reciprocal of the synaptic time constant, are close to the actual values, however show a small bias (with a mean bias of 2.5~\%). The estimates of the connectivity kernel parameters ($\theta_0$, $\theta_1$, $\theta_2$) form distributions that are approximately Gaussian. The mean parameter estimates are in good accordance with the actual values, with biases of 2.778, 7.138 and 10.2~\% for $\theta_0$, $\theta_1$ and $\theta_2$ respectively. The ratios of the estimates ($\hat\theta_0/\hat\theta_1$ and $\hat\theta_0/\hat\theta_2$) are shown in Figure~\ref{fig:ParametersRatio}. The distribution of estimates for $\hat\theta_0/\hat\theta_1$ are relativity tighter\parham{(it's 0.0812 (a bit tighter) and 3.757 respectively)}, though indicate a bias... This indicates the balance of excitation and inhibition can be estimated more accurately than the individual parameters.

\begin{table}
\centering
\begin{tabular}{cccc}
	\hline\hline Parameter & Actual Value & Mean Estimate $\pm$ Standard Deviation \\
	\hline\hline
 	$\xi$ & 0.9 & 0.925 $\pm$ 0.004 \\
	$\theta_0$ & 10 & 10.295 $\pm$ 2.879 \\
	$\theta_1$ & -8 & -8.571 $\pm$ 1.996 \\
	$\theta_2$ & 0.5 & 0.551 $\pm$ 0.083 \\
\end{tabular}
\caption{Actual parameter values and mean parameter estimates and standard deviation.}
\label{tab:Parameters estimates}
\end{table}

The reconstructed connectivity kernel from the estimated parameters can be seen in Figure~\ref{fig:KernelEstimates}. This Figure verifies the expected net shape of the estimated connectivity kernel is in good accordance to the true kernel. The shape is generally well preserved.

Figure~\ref{fig:ParametersConvergence} shows the rate of convergence for parameter estimates.
\begin{figure}
    \centering
    \subfigure[][]{\includegraphics{./Graph/zeta.pdf}}
    \subfigure[][]{\includegraphics{./Graph/theta0.pdf}}\\
    \subfigure[][]{\includegraphics{./Graph/theta1.pdf}}
    \subfigure[][]{\includegraphics{./Graph/theta2.pdf}}
\caption{Distribution of the parameter estimates over 100
realisations. The dotted line shows the location of the true parameter in each case. \textbf{a.} The distribution of $\xi$ estimates where $\xi=1-\zeta T_s $ and $\zeta$ is the synaptic time constant (true parameter: $\xi=0.9$). \textbf{b.} The distribution of the central excitatory connectivity kernel basis function parameter estimates (true parameter: $\theta_0 = 10$). \textbf{c.} The distribution of the surround inhibition connectivity kernel basis function parameter estimates (true parameter: $\theta_1 = -8$). \textbf{d.} The distribution of the longer range excitatory connectivity kernel basis function parameter estimates (true parameter $\theta_3 = 0.5$).}
\label{fig:Parameters}
\end{figure}
\begin{figure}
    \centering
\subfigure[][]{\includegraphics{./Graph/theta0_theta1_ratio.pdf}}
\subfigure[][]{\includegraphics{./Graph/theta0_theta2_ratio.pdf}}\\
\caption{Distribution of the connectivity kernel basis function parameter ratios over 100 realisations. \textbf{a.} The absolute value of the ratio of the local excitatory parameter to the surround inhibition parameter (true ratio is 1.25). \textbf{b.} The ratio of the local excitatory parameter to the longer range excitatory parameter (true ratio is 20). The dotted line shows the location of the true ratio in each case.}
\label{fig:ParametersRatio}
\end{figure}
\begin{figure}
    \centering
\subfigure[][]{\includegraphics{./Graph/KernelEstimate.pdf}}
\subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT.pdf}}\\
\caption{Results for connectivity kernel estimate. In all the subplots the true kernel is shown by the solid line, the mean kernel estimate (over 100 realisations) is shown by the dotted line and the 95~\% confidence interval is shown by the shaded grey region. \textbf{a.} Comparison of kernel estimates to true kernel in the spatial domain. \textbf{b.} Comparison of the true kernel to the estimated kernel in the frequency domain.}
\label{fig:KernelEstimates}
\end{figure}
\begin{figure}
        \centering
\includegraphics{./Graph/convergence.pdf}
\caption{The mean of the absolute error between true and estimated parameters averaged over 100 runs. All parameters converge after six iterations. The mean difference between errors in consecutive iterations falls below $10^{-4}$ after 6 iterations. }
\label{fig:ParametersConvergence}
\end{figure}

The accuracy for state estimates were evaluated by comparing the field reconstruction (see equation~\ref{DefFieldDecomp}) to the true field using the mean (over time) of the root mean square error (MRMSE) (over space) and standard deviation of the RMSE. The RMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}. A snapshot of the true field, the estimated field and their corresponding error is illustrated in Figure~\ref{fig:FieldEstimate}. This figure highlights the loss of the high spatial frequency information in the basis function decomposition.
  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region). The error remains consistent over time.} 
\label{fig:RMSE}
   \end{figure}

\begin{figure}
\centering 
\subfigure[][]{\label{fig:TrueField}\includegraphics{./Graph/TrueField90.pdf}}
\subfigure[][]{\includegraphics{./Graph/EstimatedField90.pdf}}
\subfigure[][]{\includegraphics{./Graph/EstimationError90.pdf}}\\
\caption{Examples of the neural fields. \textbf{a.} The true field from the full neural field model which was used to generate data. \textbf{b.} The reconstructed field of the reduced model, showing the effect of the basis function decomposition on the frequency content of the estimated field. \textbf{c.} The absolute error between the true and reconstructed fields, the error shows high frequency behaviour.}
\label{fig:FieldEstimate}
\end{figure}

>>>>

Given the basis function separation, the basis function width parameter was set to $\sigma_{\phi}^2=2.5$~mm. This allowed a smooth representation of the field by the basis function decomposition up to a spatial frequency $\mathbf{\nu_{c\phi}} \approx 0.12$~Hz, as described by equation~\ref{eq:CutoffFromBasisFuncWidth}. The band-limiting effect is due to the spatial frequency response of the basis functions as seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
\begin{figure} 
	\begin{center}
		\subfigure[][]{\label{fig:ObservationFreqAnal}\includegraphics{./Graph/ObservationFrequencyResponse.pdf}}\\	
		\subfigure[][]{\label{fig:SensorFunctionFreqResponse}\includegraphics{./Graph/SensorChar.pdf}}
		\subfigure[][]{\label{fig:BasisFunctionFreqResponse}\includegraphics{./Graph/BasisChar.pdf}}
	\end{center}
	\caption{\dean{Can we use $\nu_{cy}$ and $\nu_{c\phi}$ in the labels on the graphs?} Spatial frequency analysis. The dashed line shows the cut-off frequency (-3 dB point). \textbf{a.} Two-dimensional spatial frequency in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the 1-dimensional representation of the spectral characteristics. \textbf{b.} The spatial frequency response of a 1-dimensional model sensor. \textbf{c.} The spatial frequency response of a 1-dimensional neural field basis function, illustrating a narrower band width than the sensors.} 
\end{figure}

\subsection{State and Parameter Estimates} 
\label{sec:state_and_param_results}
To demonstrate the performance of the parameter and state estimation, a Monte Carlo approach was employed where $100$ realisations of the data were generated. Each realisation consisted of $500$~ms of data (sampled at $1$~kHz) and the estimation was applied to the final $400$~ms, allowing the model's initial transients to die out. The initial state and parameters were unknown to the estimator. Eleven iterations of the estimation algorithm were used to ensure the parameters  converge. Figure~\ref{fig:ParametersConvergence} shows the rate of convergence, demonstrating the parameters converged at six iterations of the estimation algorithm.
\begin{figure}
        \centering
\includegraphics{./Graph/convergence.pdf}
\caption{The mean of the absolute error between true and estimated parameters averaged over 100 runs. All parameters converge after six iterations, where the mean difference between errors in consecutive iterations falls below $10^{-4}$.}
\label{fig:ParametersConvergence}
\end{figure}

The distribution of the estimated parameters is shown in Figure~\ref{fig:Parameters}. The estimates of the parameters form distributions that are approximately Gaussian. The mean estimates, corresponding standard deviations and biases are given in Table~\ref{tab:Parameters estimates}. Although the distributions are quite wide, the mean estimates are in good concordance with the actual values. It should be noted that when the excitatory  bases of the connectivity are under (over) estimated, the inhibitory basis is typically under (over) estimated as well. This is demonstrated in Figure~\ref{fig:ParametersRatio} where we show the distribution of the ratios of the parameter estimates. 
\begin{table}
\centering
\begin{tabular}{cccc}
	\hline\hline Parameter & Actual Value & Mean Estimate $\pm$ Std. Dev. & Bias (\%) \\
	\hline\hline
 	$\xi$ & $0.9$ & $0.925\pm 0.004$ & $2.778$ \\
	$\theta_0$ & $10$ & $10.295\pm2.879$ & $2.95$\\
	$\theta_1$ & $-8$ & $-8.571\pm1.996$ & $-7.137$\\
	$\theta_2$ & $0.5$ & $0.551\pm0.083$ & $10.2$\\
	$\hat\theta_0/\hat\theta_1$ & $1.25$ & $?\pm0.0812$ &\\
	$\hat\theta_0/\hat\theta_2$ & $20$ & $?\pm3.757$ & \\
\hline\hline
\end{tabular}
\caption{Actual parameter values compared to parameter estimates.}
\label{tab:Parameters estimates}
\end{table}
\begin{figure}
    \centering
    \subfigure[][]{\includegraphics{./Graph/zeta.pdf}}
    \subfigure[][]{\includegraphics{./Graph/theta0.pdf}}\\
    \subfigure[][]{\includegraphics{./Graph/theta1.pdf}}
    \subfigure[][]{\includegraphics{./Graph/theta2.pdf}}
\caption{Distribution of the parameter estimates over 100
realisations. The dotted line shows the actual value of the parameters and the solid line shows the mean of the estimated parameters. \textbf{a.} The distribution of the $\xi$ estimates where $\xi=1-\zeta T_s $ and $\zeta$ is the reciprocal of the synaptic time constant. \textbf{b.} The distribution of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. \textbf{c.} The distribution of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$. \textbf{d.} The distribution of the longer range excitatory connectivity kernel basis function parameter estimates, $\theta_2$.}
\label{fig:Parameters}
\begin{figure}
    \centering
\subfigure[][]{\includegraphics{./Graph/theta0_theta1_ratio.pdf}}
\subfigure[][]{\includegraphics{./Graph/theta0_theta2_ratio.pdf}}\\
\caption{\dean{need to put $\left| \right|$ around $\hat\theta_0/\hat\theta_1$ in xlabel in \textbf{a.}} Distribution of the connectivity kernel basis function parameter ratios over 100 realisations. \textbf{a.} The absolute value of the ratio of the local excitatory parameter to the surround inhibition parameter (true ratio is 1.25). \textbf{b.} The ratio of the local excitatory parameter to the longer range excitatory parameter (true ratio is 20). The dotted line shows the location of the true ratio in each case.}
\label{fig:ParametersRatio}
The reconstructed connectivity kernel from the estimated parameters can be seen in Figure~\ref{fig:KernelEstimates} with the $95\%$ confidence interval. This Figure verifies the expected net shape of the estimated connectivity kernel is in good accordance to the true kernel.
\begin{figure}
    \centering
% \subfigure[][]{
\includegraphics{./Graph/KernelEstimate.pdf}%}
% \subfigure[][]{\includegraphics{./Graph/KernelEstimateFFT.pdf}}\\
\caption{Connectivity kernel estimate. The actual kernel is shown by the solid line, the mean kernel estimate (over 100 realisations) is shown by the dotted line and the 95~\% confidence interval is shown by the shaded grey region.}
\label{fig:KernelEstimates}
The accuracy for state estimates were evaluated by comparing the field reconstruction (see equation~\ref{DefFieldDecomp}) to the true field using the mean (over time) of the root mean square error (MRMSE) (over space) and standard deviation of the RMSE. The RMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}. A snapshot of the true field, the estimated field and their corresponding error is illustrated in Figure~\ref{fig:FieldEstimate}. This figure highlights the loss of the high spatial frequency information in the basis function decomposition.
  \begin{figure}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\caption{The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region). The error remains consistent over time.} 
\label{fig:RMSE}

\begin{figure}
\centering 
\subfigure[][]{\label{fig:TrueField}\includegraphics{./Graph/TrueField90.pdf}}
\subfigure[][]{\includegraphics{./Graph/EstimatedField90.pdf}}
\subfigure[][]{\includegraphics{./Graph/EstimationError90.pdf}}\\
\caption{Examples of the actual and estimated neural fields. \textbf{a.} The actual field at a single time instant from the full neural field model that was used to generate data. \textbf{b.} The reconstructed field of the reduced model, showing the effect of the basis function decomposition on the frequency content of the estimated field. \textbf{c.} The absolute error between the true and reconstructed fields. The error shows high frequency behaviour.}
\label{fig:FieldEstimate}
\end{figure}
